Application of Network Flow and Linear Programming to Scheduling of Clinicians

### 2019-08-19 paper/paper.discussion.tex
Work on discussion: split into multiple paragraphs

In this paper our goal was to develop a rather simple, yet flexible, integer linear programming formulation to generate schedules for clinical consultation (?) departments at hospitals. The difficulty in applying a linear programming approach to this task lies in the fact that ILP is an NP-hard problem, and as such the time to find an optimal solution grows exponentially as we increase the size of the problem. As a result, the vast majority of approaches taken to create schedules for similar scenarios tend to use heuristics in order to find an approximately optimal solution in a shorter time. However, even with the limitations of time complexity we are able to use the tool to generate schedules for a wide variety of simulated departments, supporting upwards of 50 total clinicians. Moreover, our formulation is able to accommodate the requests of clinicians without negatively affecting the runtime. Comparing the generated schedule to a manually created schedule in a real clinical department at a hospital, we see adherence to all required constraints as well as improved fulfillment of clinician requests, which are sometimes overlooked when creating the schedule by hand. The scenarios when the ILP formulation starts to have trouble are ones with multiple services offered within a single clinical department. Once there are more than 10 clinicians in total, providing services in 2 or more divisions, the constraint space becomes very complex, making it difficult for the tool to find an optimal solution in a reasonable amount of time. For these cases, we recommend relaxing the `No Consecutive Blocks` constraint, to simplify the problem and find a solution much faster.

>>>>

In this paper our goal was to develop a rather simple, yet flexible, integer linear programming formulation to generate schedules for clinical consultation (?) departments at hospitals. The difficulty in applying a linear programming approach to this task lies in the fact that ILP is an NP-hard problem, and as such the time to find an optimal solution grows exponentially as we increase the size of the problem. As a result, the vast majority of approaches taken to create schedules for similar scenarios tend to use heuristics in order to find an approximately optimal solution in a shorter time. \\

Our formulation includes hard constraints to ensure the schedule satisfies hospital and logistics requirements. It also aims to satisfy the work preferences of clinicians in a clinical department by optimizing a multi-goal objective function. We are able to use the tool to generate schedules for a wide variety of simulated departments, supporting upwards of 50 total clinicians. Moreover, our formulation is able to accommodate the requests of clinicians without negatively affecting the runtime. The scenarios when the ILP formulation starts to have trouble are ones with multiple services offered within a single clinical department. Once there are more than 10 clinicians in total, providing services in 2 or more divisions, the constraint space becomes very complex, making it difficult for the tool to find an optimal solution in a reasonable amount of time. For these cases, we recommend allowing consecutive blocks, to simplify the problem and find a solution much faster. \\

Our tool is also employed at the Infectious Diseases Department at St. Michael's Hospital. It provides a simple and easy-to-use user interface that can be used by any hospital staff to generate schedules well in advance. We compare the manually created schedule at St. Michael's Hospital to the optimal schedule generated with the help of our tool. We see adherence to all required hard constraints in the generated schedule, while some of them were overlooked in the manual schedule. Moreover, we see a much better fulfillment of clinician requests than in the manual schedule. This reinforces the idea that it is vital to employ automated tools when generating schedules in hospital departments, to reduce human error, balance the work-load of clinicians and improve the service provided to patients.

### 2019-08-19 paper/paper.results.tex
Work on discussion: split into multiple paragraphs

Lastly, we analyze the effect of increasing the number of divisions on the runtime of the program. Table \ref{tbl:runtime-divisions-comparison} presents the runtime for $D = \{1, 2, 3\}$ divisions and $C = \{10, 20, 30, 50\}$ clinicians in total across all divisions. For 2 divisions, a roster of more than 20 clinicians becomes impractical to solve, as searching requires more than 24 hours. For 3 divisions we can see issues with runtime occur earlier, where configurations of more than 10 clinicians are unable to be solved in 24 hours. When relaxing the consecutive blocks constraint, we see a great improvement in runtime for 2 and 3 divisions, and even a schedule for 50 clinicians or more can be generated.

>>>>

Lastly, we analyze the effect of increasing the number of divisions on the runtime of the program. Table \ref{tbl:runtime-divisions-comparison} presents the runtime for $D = \{1, 2, 3\}$ divisions and $C = \{10, 20, 30, 50\}$ clinicians in total across all divisions. For 2 or more divisions, a roster of more than 10 clinicians becomes impractical to schedule, as searching requires more than 24 hours. However, when relaxing the `No Consecutive Blocks' constraint, we see a great improvement in runtime for 2 and 3 divisions, and we are able to generate a schedule with upwards of 50 clinicians.

### 2019-08-19 paper/tbl/runtime_divisions.tex
Fix runtime table + start discussion section
Work on discussion: split into multiple paragraphs

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|c|}
		\hline
		&       \multicolumn{4}{c|}{Number of Divisions}        \\ \hline
		Number of Clinicians &  1   &   2    & 2 (Consecutive Blocks Allowed) &  3   \\ \hline
		10          & 0.22 &  0.51  &              0.26              & 1.49 \\ \hline
		20          & 0.42 & 139.42 &              0.50              &     \\ \hline
		30          & 0.63 &   -    &              0.90              &     \\ \hline
		50          & 1.06 &   -    &              1.30              &     \\ \hline
	\end{tabular}
	\caption{Comparison of program runtime (in seconds) for different number of divisions and total clinicians. ``-'' indicates that a solution was not found after 24 hours.}
	\label{tbl:runtime-divisions-comparison}%
\end{table}

>>>>

\begin{table}[htbp]
	\centering
	\begin{tabular}{|c|c||c|c||c|c|}
		\hline
		                                      & \multicolumn{5}{c|}{Number of Divisions} \\ \hline
		\makecell[l]{Number of \\ Clinicians} &  1   &  2   & 2 (CBA) &  3   &  3 (CBA)  \\ \hline
		                 10                   & 0.19 & 0.95 &  0.21   & 1.90 &   0.31    \\ \hline
		                 20                   & 0.31 &  -   &  0.41   &  -   &   0.56    \\ \hline
		                 30                   & 0.57 &  -   &  0.68   &  -   &   0.90    \\ \hline
		                 50                   & 0.78 &  -   &  1.09   &  -   &   1.66    \\ \hline
	\end{tabular}
	\caption{Comparison of program runtime (in seconds) for different number of divisions and total clinicians. `-' indicates that a solution was not found after 24 hours. `CBA' indicates that \underline{c}onsecutive \underline{b}locks were \underline{a}llowed when generating these schedules.}
	\label{tbl:runtime-divisions-comparison}%
\end{table}
